---
title: "Project Sweet"
subtitle: "How glucose test can help everybody to stable energy rate in his body"
author: "Yoni G’etahun"
date: "2022-11-11"
output: pdf_document
geometry: "left=2.5cm,right=2.5cm,top=2cm,bottom=2cm"
editor_options: 
  markdown: 
    wrap: sentence
---

## Intro

Lately I have been reading Prof Segal and prof Elinav's book about [personalized nutrition](http://personalnutrition.org/AboutGuests.aspx).

I discovered that beyond the general nutritional laws that apply to the majority of the population, there are personal laws.
These laws depend, among other things, on our lifestyle, our genetics and the microbiome in our body.

Therefore, on the recommendation of their book, I set out on a journey to discover the best diet for me.
This is the travel diary of my experience using sugar measurements as a tool for better health.

Although this is a personal experiment, you can do it on yourself and with this code create a tool your glucose result analyze.

### Goal

In this project, I had a special self goal: Better nutrition for myself.using the method from the book, I decided to take a glucose test and test my glucose and see What meals are good for me, given my lifestyle.

### Method

For 2 weeks, I have been using a glucose meter kit of 'FreeStyle' for testing my meals (including some snacks), and also my morning and night result.
I recorded using sheets my meals, tests, sleep and walk and create a small data set.

Ideal meal check has 5 tests: One with the first bite, and 4 later with half hour gap. later, in order to save test and focus on the major points I started to test 2 test after each meal with 1 hour gap. all along I did test before and after my sleep.

Then, I tried to find any correlation or effect which needed more data (a quite similar meal).

This is a BI project, not a statistical test. The Data is too small, inconsistent and non well measured for a formal experiment.

## Structure

```{r library, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
library(scales)
library(cowplot) # background_grid()
library(tidymodels)
library(lubridate) # for converting time
```
```{r import data , echo=FALSE}
sweets<- read_xlsx(
  'data/sweets.xls') %>% as.data.frame() %>% 
  mutate(Time_after_meal = hour(Time_after_meal) + minute(Time_after_meal)/60,  #convert time after to numeric for polynomial effect
         Time = as_datetime(Time) %>% 
           hms:: as_hms(),
         Glucose_result= as.numeric(Glucose_result)
         ) %>%  rename(Glucose  =Glucose_result)

day_score<- read_xlsx(
  'data/day_score.xls') %>% 
  filter(!is.na(Walk_today)) %>% 
  mutate(Sleep_before = hour(Sleep_before) + minute(Sleep_before)/60)  #convert time after to numeric for polynomial effect

main_df<- left_join(sweets, day_score, by= c("day_list_date"= "Date") )
```
```{r Functions for Ingredients split, echo=FALSE}
mini_color<- function(x)
{x<- x %>% 
    str_to_lower() %>%
    str_remove_all("white") %>% 
    str_remove_all("fake ") %>%
    str_remove_all("balls") %>%
    str_remove_all("tahun") %>%
    str_replace_all("tuna", "fish") %>%
    str_replace_all(" ,", ",") %>% 
    str_replace_all(', ', ",") %>% 
    str_squish() %>%
    {ifelse(str_detect(.,'NA'), NA,.)} %>%
    str_split(",") %>%
    map(~ paste0("has_", .) ) %>%
    map(~str_replace_all(.," ", "_"))
  x}

close_em_col<- function(x) #wider the colour
{
  cbind(x, x$Ingredients %>% 
  qdapTools::mtabulate()) %>%
  select(-c(Ingredients))
}

Ingredients_df<- main_df%>%
  mutate(Ingredients= mini_color(Ingredients)) %>%
  close_em_col() %>% as.data.frame() %>%  select(-has_cruasont)
```

The data is made of 2 data sheets I made:

- **sweets**: data per test

- **day_score**: data per day ingredients

Later you will see:

- **main_df**: combined data before splitting the ingredients

- **Ingredients_df**: combined data 'before splitting the ingredients with split column by ingredients

### Average day

Here we can see an average day.
From this graph, I can choose which meal and day to look for looking for a good or bad example

```{r over the day data, message=FALSE, warning=FALSE}
sweets %>%
  filter(!is.null(Glucose), !Dinner %in% c('tzom', 'sleep'),) %>% 
  mutate(Date = format(Date, "%m-%d")) %>% 
  mutate(Date= as.factor(Date)) %>% 
  ggplot(aes(x= Time, y= Glucose, color= Date))+
  geom_line(size= 1.1)+ geom_point(size= 0.9, shape= 1, color= "Black")+
  facet_wrap(~Date,3)+
  guides(color=guide_legend(title="Day"))+ xlab("Time")+
  background_grid(minor = "yx", color.major= "gray50")+
    theme(legend.position = "none")+ggtitle("Glucose result by day")

sweets %>%
  filter(!is.null(Glucose),!Date %in% c('2022-10-22','2022-11-04')) %>% 
  mutate(Date = format(Date, "%m-%d")) %>% 
  mutate(Date= as.factor(Date),
         Glucose= scale(Glucose)) %>% ##,
  ggplot(aes(x= Time_after_meal, y= Glucose, color= Date))+
  geom_line(size= 1.15)+ geom_point(size= 1.1, shape= 2)+facet_wrap(~Dinner)+
  guides(color=guide_legend(title="Day"))+ xlab("H after meal")+
  background_grid(minor = "y", color.major= "gray50")+
  ggtitle("Glucose result by meal")
```

From this data, I can see which meal is my "to go" and which us a big no.
For example, worst snacks, or best breakfast:

```{r example who is biggest, message=FALSE}
main_df%>%
  filter(Dinner== 'snack', Type != 'zero') %>%
  mutate(Time_after_meal= round(Time_after_meal,2)) %>% 
  select(ID,Date, Glucose, Type, Ingredients, Time_after_meal) %>% 
  arrange(desc(Glucose)) %>% slice(1:5)
  
main_df%>%
  filter(Dinner== 'morning', Type != 'zero') %>%
  mutate(Time_after_meal= round(Time_after_meal,2)) %>% 
  select(ID,Date, Glucose, Type, Ingredients, Time_after_meal) %>% 
  arrange(Glucose) %>% slice(1:5)
```

Here I try to find a general effect of walkin/ sleaping on the glucose

```{r daily effect, message=FALSE}
main_df%>%
  mutate(active= walking_before== "T"| walking_while== "T" ) %>% 
  ggplot(aes(y= Glucose, x= Walk_today, color= Dinner))+
  geom_point(alpha= 0.8)+
  geom_smooth(aes(x= Walk_today, y= Glucose), se= F, method = "lm")+
  xlab("steps walked today")+ggtitle("Is it healthy to walk?",
  subtitle = "Glucose result by steps walked this day and meal")+
  scale_x_continuous(labels= scales::comma)

main_df%>%
  mutate(active= walking_before== "T"| walking_while== "T" ) %>% 
  ggplot(aes(y= Glucose, x= Walke_d_before, color= Dinner))+
  geom_point(alpha= 0.8)+
  geom_smooth( se= F, method = "lm")+xlab("steps walked a day before")+
  ggtitle("Is it healthy to walk ahead?",
  subtitle = "Glucose result by steps walked the other day and meal")+
  scale_color_brewer(palette="Set1")+scale_x_continuous(labels= scales::comma)
```

as far as I can see with no model creations, there is no corellation.

## Ingredients effect

### Daily effect

```{r combine data, warning=FALSE, message=FALSE}
Ingredients_df %>%  select(c(1:7,9:17)) %>% sample_n(5)

main_df %>%
  filter(Dinner %in% c('sleep', 'tzom')) %>% 
  ggplot(aes(y= Glucose, x= Sleep_before, shape= Work==1,color= Dinner))+
  geom_point()+geom_smooth(method = "lm", se= F)+
  theme(legend.position="bottom"
)

main_df %>%
  filter(Dinner %in% c('sleep', 'tzom')) %>% 
  ggplot(aes(y= Glucose, x= Date, color= Dinner, shape= Glucose<= 1.5))+
  geom_point()+geom_smooth(se= F, method = "lm")+
theme(legend.position="bottom"
)
```

I can see a clue that sleeping better might help reduce morning glucose rate, and also that in general, my study arte of reducing night glucose rate is fine, even removing one elevating point.

```{r Is sugar that bad, warning=FALSE, message=FALSE}
Ingredients_df %>%
  mutate(sugar= has_ice_coffe |has_5_ckookies|has_waffels| has_ice_cream| has_jam| 
           has_banana|has_apple| has_shalva| has_snack)%>% 
  ggplot(aes(y= Glucose, x=Walk_today, color= sugar))+
  geom_point()+
  geom_smooth(alpha= 0.2, method = "lm", se= F)+
  theme(panel.grid.major = element_line(color = "gray40",
                                  size = 0.5,
                                  linetype = 1),
        panel.grid.minor.x = element_line(color = "gray80",
                                  size = 0.20,
                                  linetype = 1),
        legend.position="bottom"
        )+scale_x_continuous(labels= scales::comma)+
  ggtitle("Is sugar that bad?",subtitle ="Glucose rate by steps walked this day, split by sweet snakcs and general food")
```

### Meal effect

We can also create example of compering lunch with and without cauliflower and broccoli:

```{r}
Ingredients_df %>%
  filter(Dinner %in% 'lunch') %>% 
  mutate(fisher= has_fish| has_salmon,
         flower= has_cauliflower | has_broccoli) %>% 
  ggplot(aes(y= Glucose, x= as.factor(flower),color= (flower==1)))+
  geom_point()+geom_boxplot(alpha= 0.4, fill= "Gray60")+
  ggtitle("Does cauliflower and broccoli good for me?", subtitle = "Compare lunch with or whithout broccoli and cauliflower")+xlab("")+
  scale_color_brewer(palette="Accent")
```

In the following 4 graphs, you will see with me how different ingredients contained in different meal's glucose result:

```{r sweet compare, echo=FALSE}
sweet_list_0<- c('has_ice_coffe' ,'has_5_ckookies','has_waffels' ,'has_ice_cream', 'has_jam', 'has_banana','has_apple',
                 'has_shalva','has_snack', 'has_chocolate')
sweet_list<- str_remove_all(sweet_list_0,"has_")

Ingredients_df %>%
  mutate(sugar= has_ice_coffe |has_5_ckookies|has_waffels| has_ice_cream| has_jam| has_banana|has_apple| has_shalva| has_snack |has_chocolate,
         Glucose= scale(Glucose))%>% filter(sugar== T) %>% 
  pivot_longer(cols = starts_with("has"),
    names_to = "Sweet_type",
    values_to = "value", values_drop_na = TRUE) %>%
  select(-c(sugar)) %>% 
  mutate(Sweet_type= str_remove_all(Sweet_type,"has_")) %>% 
  filter(value==1, Sweet_type %in% sweet_list) %>% 
  
  ggplot(aes(y= Glucose, x=Sweet_type,color= Sweet_type))+
  geom_point()+#geom_smooth(se= F, method = "lm")
  geom_boxplot(alpha= 0.2)+
  theme(legend.position = "none")+
  ggtitle("Glucose by sweet ingrediang indicator") + xlab("")
```

The same goes for those graphs:

```{r carbon compare, echo=FALSE}
carbon_list_0<- c('has_quinoa' ,'has_kuskus','has_oatmeal' ,'has_pasta', 'has_patato',
                  'has_rice','has_sambusak','has_tortia','has_laffa')
carbon_list<- str_remove_all(carbon_list_0,"has_")
```

```{r carbon graph, echo=FALSE}
Ingredients_df %>%
  mutate(carbon= has_quinoa |has_kuskus |has_oatmeal |has_pasta |has_potato |
           has_rice |has_laffa| has_sambusak |has_tortia
           )%>% 
  filter(carbon== T,Type!=  "zero") %>% 
  pivot_longer(cols = starts_with("has"),
    names_to = "carbon_type",
    values_to = "value", values_drop_na = TRUE) %>%
  select(-c(carbon)) %>% 
  mutate(carbon_type= str_remove_all(carbon_type,"has_")) %>% 
  filter(value==1, carbon_type %in% carbon_list) %>% 
  
  ggplot(aes(y= Glucose, x= carbon_type,color= carbon_type, fill= "Black"))+
  geom_point()+ geom_boxplot(alpha= 0.2)+ theme(legend.position = "none")+
  ggtitle("Glucose by Carbohydrate ingrediang indicator") + xlab("")
```

```{r protein compare, echo=FALSE}
protein_list_0<- c('has_salmon' ,'has_fish','has_chicken','has_meat', 'has_peas','has_humus',
                   'has_veg_hamburge','has_falafe','has_liver','has_egg', 'has_tofu', 'has_faba_bean','has_shawarma')
protein_list<- str_remove_all(protein_list_0,"has_")
```

```{r protein graph, echo=FALSE}
Ingredients_df %>%
  mutate(protein= has_salmon |has_fish |has_chicken |has_meat |has_peas|has_humus |
           has_veg_hamburger| has_falafel| has_liver | has_egg| has_tofu|has_faba_bean|has_shawarma)%>% 
  filter(protein== T, Type!=  "zero") %>% 
  pivot_longer(cols = starts_with("has"),
    names_to = "protein_type",
    values_to = "value", values_drop_na = TRUE) %>%
  select(-c(protein)) %>% 
  mutate(protein_type= str_remove_all(protein_type,"has_")) %>% 
  filter(value==1, protein_type %in% protein_list) %>% 
  
  ggplot(aes(y= Glucose, x= protein_type,color= protein_type, fill= "Black"))+
  geom_point()+ geom_boxplot(alpha= 0.2)+
  theme(legend.position = "none")+
  ggtitle("Glucose by protein ingrediang indicator") + xlab("")
```

```{r vegetable compare, echo=FALSE}
vegetable_list_0<- c('has_potato' ,'has_pomegranate' ,'has_broccoli' ,'has_beetroot' ,'has_cauliflower',
           'has_peas' , 'has_salad', 'has_tomato' ,'has_garlic', 'has_vegetables','has_mushroom')
vegetable_list<- str_remove_all(vegetable_list_0,"has_")
  
Ingredients_df %>%
  mutate(vegetable= has_potato |has_pomegranate |has_broccoli |has_beetroot | has_cauliflower|
           has_peas |has_salad| has_tomato | has_garlic| has_vegetables|has_mushroom)%>% 
  filter(vegetable== T, Type!=  "zero") %>% 
  pivot_longer(cols = starts_with("has"),
    names_to = "vegetable_type",
    values_to = "value", values_drop_na = TRUE) %>%
  select(-c(vegetable)) %>% 
  mutate(vegetable_type= str_remove_all(vegetable_type,"has_")) %>% 
  filter(value==1, vegetable_type %in% vegetable_list) %>% 
  
  ggplot(aes(y= Glucose, x= vegetable_type,color= vegetable_type, fill= "Black"))+
  geom_point()+#geom_smooth(se= F, method = "lm")
  geom_boxplot(alpha= 0.2)+
  theme(legend.position = "none")+
  ggtitle("Glucose by vegetable ingrediang indicator") + xlab("")
```

### Ingredients linear model

Ideally, I would use a statistical model in order to understand which food is better for my balance, and create a noun recommendation.
Unfortunately, this is not a big enough data nor the variables noun recommendation enough.
Even though, I made linear regression for myself in order to have some clue for how my body works.

```{r bake data}
options(scipen = 10)

baking_food<- Ingredients_df %>%
  recipe(Glucose~.) %>% 
  update_role(ID,Date,Type,Time, new_role = "sider") %>% 
  step_rm(has_role("sider"),day_list_date) %>% 
  step_mutate(Time_after_meal2= Time_after_meal^2,
              has_fish= has_fish|has_salmon,
              has_chicken= has_chicken| has_shawarma,
              has_sweet_snack= has_5_ckookies| has_waffels) %>% 
  step_rm(has_salmon,has_vegetables, has_beetroot, has_hot,has_shawarma,has_5_ckookies,has_waffels)

baked_food<- baking_food %>% prep(Ingredients_df)%>% bake(Ingredients_df) 
lm_par <- linear_reg() %>% set_mode('regression') %>% set_engine("lm")
lm_fit <- lm_par %>% fit(Glucose ~ . , baked_food)
```

Now, I can show a pick to which foods should I eat more / less

```{r lm summery}
lm_beta<- summary(lm_fit$fit)
lm_beta
lm_beta$coefficients %>% 
  as.data.frame() %>% #colnames()
  rename(Pr_value= 4) %>% 
  filter( Pr_value<= 0.35, Estimate<= 0) %>% 
  arrange(Estimate)%>%  slice(1:7) %>% mutate(across(everything(), ~round(.,4)))%>% 
  select(-'t value')

lm_beta$coefficients %>% 
  as.data.frame() %>% #colnames()
  rename(Pr_value= 4) %>% 
  filter( Pr_value<= 0.35, Estimate>= 0,
          row.names(lm_beta$coefficients)!= '(Intercept)') %>% 
  arrange(desc(Estimate)) %>%  slice(1:6) %>% mutate(across(everything(), ~round(.,4))) %>% 
  select(-'t value')
```

In a glance, with $pr(>|t|)< 0.35$ I should avoid snacks(energy snack), sambusak, shalva and sweet snacks, and use more coffee, kuskus, yogurt and egg

```{r visual result, echo=FALSE, eval=FALSE}
lm_beta<- summary(lm_fit$fit)
x_glocos<- 0:120
y_glocos<-  Ingredients_df$Glucose %>% mean()+
  x_glocos*  lm_beta$coefficients['Time_after_meal' ,1]+
  x_glocos^2*lm_beta$coefficients['Time_after_meal2',1]
plot(x_glocos, y_glocos, type = "l", col= "Red")
```

As said before, there is some correlations that could affect this model, and therefore the liability is quite fragile.
Here are some of the corelations:

```{r GGally, warning=FALSE, message=FALSE, echo=FALSE}
#cor_food<- 
lm_beta$cov.unscaled %>% as.data.frame() %>% 
  select(starts_with("has_")) %>%
  select(-has_sambusak, -has_na,has_snyders, -has_milk, -has_honey_cake,
         -has_laffa, -has_cheese,-has_butter, -has_snyders) %>%  #remove some uncommon
  filter(str_detect(row.names(lm_beta$cov.unscaled), 'has_')) %>%
  rename_all(~str_remove_all(., "has_")) %>% 
GGally:: ggcorr(method = c("everything", "pearson"))+ #visualization of correlations
  ggtitle("Correlation graph of some common food")
```
```{r unwanted GGally, echo=FALSE, eval=FALSE}
#lm_beta$cov.unscaled%>% as.data.frame()%>% select(!starts_with("has_"))%>% 
#GGally:: ggcorr(method= c("everything","pearson"))
```

## conclusions:

After a few days in the project, I felt my energy is more balanced.
I became less tired during lunch, evening and night.

The experiment's method forced me to eat less snacks, and as a result I found out how dessertless meal makes me feel more full, didn't eat afternoon meal and were more energetic later in the way home.

Moreover, I realize adding a balanced amount of more fat and protein to my meals can stable my sugar rate and cause me to eat less calories without calculating calories or carbohydrates.

Looking for a better body energy balance, I would consider sleeping a slightly better and more might help me maintaining balanced.
A walk might help as well.

Now here are some food based self advice:

-   Adding fat yogurt to the oatmeal can balance the healthy carbohydrates with oil that stable my sugar.

-   Moderate quantity of carbohydrates like Couscous or quinoa on lunch can be fine as long as the meal contain protein & some fat

-   A fish is a good protein to add to my meals which I hadn't thought on till now

-   Ice cream is the best snack for me, this is might due to the fat and water covering the sugar.
    Nuts with fresh fruit are also excellent, and Ice coffee is ok.
    

-   Broccoli and cauliflower are suspected of being good for lunch balance

-   When I have a bad meal, coffee can lower my glucose, and this might relevant to tea as well

-   The last 2 hours before sleep shouldn't contain too big meals (no sambusak before bed...)

**Note!**

This is a self experiment.

There is some path that might be good for most people like mix fat in carbohydrates and space meals,
but some might not. This self diary can help you create your own diary!

Hope for better health and energy balance for all of you!
